---
title: Spatially varying catchability simulation study results
author: John Best
date: "`r Sys.time()`"
---

```{r echo=FALSE,messages=FALSE,warnings=FALSE}
library(tidyverse)
eval_dir <- normalizePath("evaluation")
knitr::opts_chunk$set(echo = FALSE)

studies <- c("qdevscaling",
             "sharedq",
             "prefintensity",
             "densdepq")
estmods <- c("model",
             "survey",
             "spatial_ab",
             "spatial_q")
```

# Operating model descriptions

- **Catchability deviation scaling**: Varies the magnitude of spatial deviations
  in catchability
- **Shared catchabilty deviations**: Survey vessels are also affected by
  catchability deviations, with the intensity scaled as a fraction of the
  deviation experienced by the commercial fleet
- **Preference intensity**: The commercial fleet chooses fishing locations based
  on the product of abundance and catchability raised to a power
- **Density-dependent catchability**: Catchability is a function of abundance;
  catchability increases linearly with abundance.
  
# Estimation model descriptions

- **`model`**: Simple, non-spatial GLM
- **`survey`**: Uses only survey data, includes a spatial random effect
- **`spatial_ab`**: Uses survey data plus 4,000 commercial observations,
  includes an abundance spatial random effect
- **`spatial_q`**: Uses survey data plus 4,000 commercial observations, includes
  both abundance and catchability spatial random effects

# Successful model fits

The following tables list the number of model fits that are used in the
following figures. The goal is 100 replicates for each, but this takes multiple
days of computation time, so this is the set currently complete.

## Catchability deviation scaling

```{r}
read_csv(file.path(eval_dir, studies[1], "pdhess_wide.csv"),
         col_types = cols(estmod = col_factor(levels = estmods),
                          .default = col_integer())) %>%
  arrange(estmod) %>%
  knitr::kable(.)
```

## Shared catchability deviations

```{r}
read_csv(file.path(eval_dir, studies[2], "pdhess_wide.csv"),
         col_types = cols(estmod = col_factor(levels = estmods),
                          .default = col_integer())) %>%
  arrange(estmod) %>%
  knitr::kable(.)
```

## Preference intensity

```{r}
read_csv(file.path(eval_dir, studies[3], "pdhess_wide.csv"),
         col_types = cols(estmod = col_factor(levels = estmods),
                          .default = col_integer())) %>%
  arrange(estmod) %>%
  knitr::kable(.)
```

## Density-dependent catchability

```{r}
read_csv(file.path(eval_dir, studies[4], "pdhess_wide.csv"),
         col_types = cols(estmod = col_factor(levels = estmods),
                          .default = col_integer())) %>%
  arrange(estmod) %>%
  knitr::kable(.)
```

# RMSE

## Catchability deviation scaling

```{r}
knitr::include_graphics(file.path(eval_dir, studies[1], "rmse_plot.svg"))
```

## Shared catchability deviations

```{r}
knitr::include_graphics(file.path(eval_dir, studies[2], "rmse_plot.svg"))
```

## Preference intensity

```{r}
knitr::include_graphics(file.path(eval_dir, studies[3], "rmse_plot.svg"))
```

## Density-dependent catchability

```{r}
knitr::include_graphics(file.path(eval_dir, studies[4], "rmse_plot.svg"))
```

# Bias

## Catchability deviation scaling

```{r}
knitr::include_graphics(file.path(eval_dir, studies[1], "bias2_plot.svg"))
```

## Shared catchability deviations

```{r}
knitr::include_graphics(file.path(eval_dir, studies[2], "bias2_plot.svg"))
```

## Preference intensity

```{r}
knitr::include_graphics(file.path(eval_dir, studies[3], "bias2_plot.svg"))
```

## Density-dependent catchability

```{r}
knitr::include_graphics(file.path(eval_dir, studies[4], "bias2_plot.svg"))
```
