---
title: Update on Habitat-based catchability
author: John Best
date: September 22, 2021
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(hdf5r)
```

```{r}
study <- "habq"
repl <- 1
opmod <- 6

prep_path <- sim_file_paths(study, repl, opmod)$prep
h5 <- h5file(prep_path, mode = "r")
rocky_hab <- h5[["rocky_hab"]]$read()
```

```{r}
study <- "habq"
repl <- 1
opmods <- 1:6

init_pop <- lapply(opmods,
                   function(om)
                     read_popstate(study, repl, om, filetype = "h5")[, , 1])

png("habq/init_pop.png", width = 768)
plot_popstate(init_pop, plot_cols = 3)
dev.off()
```

```{julia}
using SpatQSims
using Plots

spec = HabQSpec(1, 1)
hab = get_habitat(spec)

heatmap(hab[2]; aspect_ratio = 1)
```

# manuscript

## motivation

Consistent sampling procedures in fishery-independent surveys allow a reasonable
assumption of constant catchability over time so that indices of abundance can
be derived from their catches. However, this limits the ability of the
fishery-independent survey to adopt newer technologies and techniques that
increase catchability. Fishery-dependent vessels are not constrained in this
way, and often actively develop fishing methods that increase their
catchability. This mismatch can reduce industry confidence in the accuracy of
fishery-independent surveys (ref). One example of this is the ability of
fishery-dependent vessels to trawl closer to or within areas of rocky habitat
that the survey will not sample at all. If areas that are more difficult for the
fishery-independent survey to sample are areas of higher abundance, the
fishery-independent survey may miss important changes in abundance. This is
particularly true if the fishery-dependent vessels are able to harvest fish from
these areas. In this case, including fishery-dependent catches in an index of
abundance can fill in the gaps of the fishery-independent survey.

## operating model

The next simulation study addresses the case where habitat is associated with
both movement preference and catchability, and catchability differs between the
fishery-independent survey and fishery-dependent vessels. For each of the 100
replicates, a habitat covariate is generated from a Gaussian random field. The
generated field is then thresholded so that 25% of the area is designated as
"rocky" habitat. A movement preference for the rocky habitat relative to the
non-rocky habitat is assigned a value between 0.3 (preference for non-rocky
habitat) and 5 (strong preference for rocky habitat). A relative preference of 1
is included in this range, indicating the case where there is no movement
preference between the two habitat types. Catchability in the non-rocky habitat
is 0.01 for both the fishery-independent survey and fishery-dependent vessels.
Within the rocky habitat catchability for the fishery-dependent vessels is
reduced to 0.001, while catchability for the fishery-dependent vessels is only
reduced to 0.009. This represents the technology and technique differences
between the two vessel types. From there, the simulation proceeds as with the
other simulation studies.

## results

When fishery-independent surveys are subject to spatially varying catchability,
indices of abundance derived from their catches can be hyperstable even in the
absence of spatially varying abundance. Hyperstability increases with the fish's
preference for the lower-catchability areas. Much of this hyperstability can be
eliminated by including catches that are not subject to the same cathability
reduction, and further still by including spatially varying catchability in the
estimation model (Figure X).

## discussion

Fishery-dependent catches may contain important information that is not
available from fishery-independent surveys, particularly in situations where the
fishery-dependent vessels can fish areas marginally accessible to the
fishery-independent survey. When these differences manifest as differences in
catchability, including a spatial random effect to account for differences in
catchability can further stabilize estimated indices of abundance.

This technique could be extended by allowing temporal variation of the spatial
catchability component of the estimation model. This would allow for a situation
where the fishery-independent survey and the fishery-dependent vessels start
with the same spatial catchability structure, but over time the
fishery-dependent vessels develop techniques and technology that increases their
catchability in particular areas of the fishery.
