---
title: Operating model example
output:
  html_document:
    toc: true
    toc_depth: 1
params:
    study: qdevscaling
    repl: 1
    opmod: 1
    max_T: 15
    root_dir: ..
---

```{r include = FALSE}
library(spatq)
library(tidyverse)
library(patchwork)
library(hdf5r)

knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)

om_files <- sim_file_paths(params$study,
                           params$repl,
                           params$opmod,
                           params$root_dir)
```

```{r}
source("om_sims_funs.R")

## Spatial binning for e.g. effort plots
brks <- seq(0, 100, length = 21)
sz <- brks[2] - brks[1]
cntrs <- brks[1:20] + diff(brks) / 2

pars <- om_pars(params$study, params$opmod)
```

## OM parameter values

This was part of the simulation study addressing the effect of the `r switch(params$study, qdevscaling = "magnitude of catchability variation", sharedq = "magnitude of catchability variation for the survey relative to the commercial fleet", prefintensity = "intensity of preferential sampling")`.

- Base catchability: $`r pars$base_q`$
- Marginal SD of commercial log-catchability deviations: $`r pars$logq_sd`$
- Marginal SD of survey log-catchability deviations: $`r pars$sharedq * pars$logq_sd`$
- Preference power: $`r pars$pref_power`$

# Overall trends

## Abundance

```{r}
pop_df <- read_popstate(params$study,
                        params$repl,
                        params$opmod,
                        params$root_dir) %>%
  filter(time <= params$max_T)

pop_df %>%
  ggplot(aes(x = time, y = pop)) +
  geom_line() +
  scale_x_continuous(breaks = seq(5, 15, 5),
                     minor_breaks = 1:15,
                     expand = expansion(0.05, 0)) +
  scale_y_continuous(limits = c(0, NA),
                     expand = expansion(c(0, 0.05), 0)) +
  labs(y = "Total biomass") +
  theme_minimal() +
  theme(axis.line = element_line())
```

## Catch

```{r}
catch_df <- read_catch(params$study,
                       params$repl,
                       params$opmod,
                       params$root_dir) %>%
  filter(time <= params$max_T) %>%
  mutate(vessel_idx = factor(vessel_idx,
                             labels = c("Survey", "Commercial")))
```

```{r}
catch_df %>%
  group_by(time, vessel_idx) %>%
  summarize(catch_biomass = sum(catch_biomass)) %>%
  ggplot(aes(x = time, y = catch_biomass, color = vessel_idx)) +
  geom_line() +
  facet_grid(vessel_idx ~ ., scales = "free_y") +
  scale_x_continuous(breaks = seq(5, 15, 5),
                     minor_breaks = 1:15,
                     expand = expansion(0.05, 0)) +
  scale_y_continuous(limits = c(0, NA),
                     expand = expansion(c(0, 0.05), 0)) +
  labs(y = "Catch biomass") +
  scale_color_discrete(guide = FALSE) +
  theme_minimal() +
  theme(axis.line = element_line())
```

## Naive indices of abundance

These are simply the mean catches within each year, normalized to have a geometric mean of 1.

```{r}
comm_index <- catch_df %>%
  filter(vessel_idx == "Commercial") %>%
  group_by(time) %>%
  summarize(tot_catch = sum(catch_biomass)) %>%
  mutate(vessel_idx = "Commercial",
         index = rescale_index(tot_catch)$index) %>%
  select(-tot_catch)

surv_index <- catch_df %>%
  filter(vessel_idx == "Survey") %>%
  group_by(time) %>%
  summarize(tot_catch = sum(catch_biomass)) %>%
  mutate(vessel_idx = "Survey",
         index = rescale_index(tot_catch)$index) %>%
  select(-tot_catch)

combo_index <- catch_df %>%
  group_by(time) %>%
  summarize(tot_catch = sum(catch_biomass)) %>%
  mutate(vessel_idx = "Combo",
         index = rescale_index(tot_catch)$index) %>%
  select(-tot_catch)

true_index <- pop_df %>%
  mutate(vessel_idx = "True",
         index = rescale_index(pop)$index) %>%
  select(-pop)

rbind(comm_index, surv_index, combo_index, true_index) %>%
  mutate(vessel_idx = fct_relevel(vessel_idx,
                                  "Survey",
                                  "Commercial",
                                  "Combo",
                                  "True")) %>%
  ggplot(aes(x = time, y = index, color = vessel_idx)) +
  geom_line() +
  ## geom_line(aes(x = time, y = index), data = true_index, color = "black", inherit.aes = FALSE) +
  scale_x_continuous(breaks = seq(5, 15, 5),
                     minor_breaks = 1:15,
                     expand = expansion(0.05, 0)) +
  scale_y_continuous(limits = c(0, NA),
                     expand = expansion(c(0, 0.05), 0)) +
  scale_color_manual(values = c("#F8766D", "#00BFC4", "#7CAE00", "#000000")) +
  labs(y = "Naive index") +
  theme_minimal() +
  theme(axis.line = element_line(),
        legend.title = element_blank(),
        legend.position = "bottom")
```

# Abundance

```{r}
pop_file <- h5file(om_files$poph5, mode = "r")
pop_array <- pop_file[["popstate"]]$read()
pop_file$close()
```

```{r}
spatpop_df <- cross_df(list(s1 = seq(0.5, 99.5),
                            s2 = seq(0.5, 99.5),
                            time = 1:params$max_T)) %>%
  mutate(pop = as.vector(pop_array[, , seq_len(params$max_T)]))
```

```{r}
pop_plot <- ggplot(spatpop_df, aes(x = s1, y = s2, fill = pop)) +
  geom_raster() +
  scale_fill_viridis_c(guide = FALSE) +
  facet_wrap(~ time, nrow = 3) +
  coord_equal(xlim = c(0, 100),
              ylim = c(0, 100)) +
  om_theme()
pop_plot
```

# Catch

## Survey

```{r}
survey_catch_plot <- catch_df %>%
  group_by(time, s1, s2) %>%
  filter(vessel_idx == "Survey") %>%
  summarize(tot_catch = sum(catch_biomass)) %>%
  ggplot(aes(x = s1, y = s2, color = tot_catch, size = tot_catch)) +
  ## geom_tile(width = 1, height = 1) +
  geom_point() +
  scale_color_viridis_c(option = "B") +
  guides(color = FALSE,
         size = FALSE) +
  facet_wrap(~ time, nrow = 3) +
  coord_equal(xlim = c(0, 100), ylim = c(0, 100)) +
  om_theme()
survey_catch_plot
```

## Commercial (binned 5×5)

The full 100×100 grid is too sparse and noisy when looking at the whole time series.

```{r}
comm_df <- catch_df %>%
  filter(vessel_idx == "Commercial") %>%
  mutate(g1 = cut(s1, brks, labels = FALSE),
         g2 = cut(s2, brks, labels = FALSE)) %>%
  group_by(time, g1, g2) %>%
  summarize(tot_catch = sum(catch_biomass),
            tot_effort = sum(effort)) %>%
  mutate(s1 = cntrs[g1],
         s2 = cntrs[g2])

comm_catch_plot <- ggplot(comm_df, aes(x = s1, y = s2, fill = tot_catch)) +
  geom_tile(width = sz, height = sz) +
  scale_fill_viridis_c(option = "B",
                       guide = FALSE) +
  facet_wrap(~ time, nrow = 3) +
  coord_equal(xlim = c(0, 100), ylim = c(0, 100)) +
  om_theme()
comm_catch_plot
```

# Commercial targeting

## Expected commercial effort

This is calculated as the probability of being chosen, multiplied by 2000 total
trips.

```{r}
pref_df <- spatpop_df %>%
  group_by(time) %>%
  mutate(tot_pop = sum(pop),
         target_prob = pop / tot_pop,
         expect_eff = target_prob * 2000)
  
pref_df %>%
  ggplot(aes(x = s1, y = s2, fill = expect_eff)) +
  geom_raster() +
  scale_fill_binned() +
  coord_equal(xlim = c(0, 100),
              ylim = c(0, 100)) +
  facet_wrap(~ time, nrow = 3) +
  om_theme() +
  theme(legend.title = element_blank())
```

## Realized commercial effort

```{r}
ggplot(comm_df, aes(x = s1, y = s2, fill = tot_effort)) +
  geom_tile(width = sz, height = sz) +
  scale_fill_viridis_b(option = "E") +
  facet_wrap(~ time, nrow = 3) +
  coord_equal(xlim = c(0, 100), ylim = c(0, 100)) +
  om_theme()
```

# Catchability

```{r}
## logq_sd <- om_val(params$study, params$opmod)
prepf <- h5file(file.path(params$root_dir, "prep.h5"), "r")
logq_devs <- prepf[["log_catchability_devs"]]$read()[, , params$repl]
prepf$close()
logq_df <- spatpop_df %>%
  filter(time == 1) %>%
  mutate(logq_dev = as.vector(logq_devs),
         comm_q = comm_catchability(logq_dev, pars),
         surv_q = survey_catchability(logq_dev, pars),
         comm_popq = pop * comm_q,
         surv_popq = pop * surv_q)

cb_plot <- logq_df %>%
  select(s1, s2, comm_q, surv_q) %>%
  pivot_longer(cols = c(comm_q, surv_q),
               names_to = "fleet",
               values_to = "q") %>%
  mutate(fleet = ifelse(fleet == "comm_q",
                        "Commercial", "Survey")) %>%
  ggplot(aes(x = s1, y = s2, fill = q)) +
  geom_raster() +
  scale_fill_viridis_c(option = "A") +
  coord_equal(xlim = c(0, 100),
              ylim = c(0, 100)) +
  facet_grid(fleet ~ .) +
  ggtitle("Catchability") +
  om_theme() +
  theme(legend.title = element_blank())
  
abq_plot <- logq_df %>%
  select(s1, s2, comm_popq, surv_popq) %>%
  pivot_longer(cols = c(comm_popq, surv_popq),
               names_to = "fleet",
               values_to = "popq") %>%
  mutate(fleet = ifelse(fleet == "comm_popq",
                        "Commercial", "Survey")) %>%
  ggplot(aes(x = s1, y = s2, fill = popq)) +
  geom_raster() +
  scale_fill_viridis_c(option = "C") +
  coord_equal(xlim = c(0, 100),
              ylim = c(0, 100)) +
  facet_grid(fleet ~ .) +
  ggtitle("Expected catch (abundance × catchability)") +
  om_theme() +
  theme(legend.title = element_blank())

cb_plot + abq_plot
```

